---
title: "Reproducible Research: Peer Assessment 1"
author: Carolina Abreu
date: October, 2015
output: 
  html_document:
    keep_md: true
---

## Prepare the R environment

Throughout this report when writing code chunks in the R markdown document, **always use echo = TRUE** so that someone else will be able to read the code. 

First, we set echo equal a **TRUE** and results equal a **'hold'** as global options for this document.  
```{r set_options}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

### Load required libraries
```{r call_libraries}
library(data.table)
library(ggplot2)
```


## Loading and preprocessing the data

### Download the data file and set the working directory
**Note**: To this script to work is necessary that the file activity.csv is in the current working directory. File can be downloaded from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

### Load libraries
```{r call_libraries}
library(data.table)
library(ggplot2)
```

### Load the data
```{r read_data}
rdata <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
```

### "Cleaning" the data

Converting **date** field to `Date` class and **interval** field to `Factor` class.

```{r tidy_data}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
```

### Glance at data

```{r check_data}
str(rdata)
```

## What is mean total number of steps taken per day?

First, we calculate the total steps per day.

```{r pre_calc_stepsperday}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

Then, make a histogram of the total number of steps taken per day.

```{r histo}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

Finaly, calculate the ***mean*** and ***median*** of the number of steps taken per day.

```{r meanmedian}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```

Answer: The mean is **`r format(steps_mean,digits = 8)`** and median is **`r format(steps_median,digits = 8)`**.

## What is the average daily activity pattern?

Calculate the aggregation of steps by intervals of 5-minutes and convert them as integers and save them in a data frame called `steps_per_interval`.

```{r steps_interval}
steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
##this helps in plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```


Make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:


```{r plot_time_series}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="orange", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```

Find the 5-minute interval with the containing the maximum number of steps:

```{r max_interval}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
```

Answer:The **`r max_interval$interval`<sup>th</sup>** interval has maximum **`r round(max_interval$steps)`** steps.


## Imputing missing values:

### Total number of missing values:

```{r tot_na_value}

missing_vals <- sum(is.na(rdata$steps))

```

Answer: The total number of ***missing values*** are **`r missing_vals`**.

### Strategy for filling in all of the missing values in the dataset

To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in our case the total median is not much far away from total mean, and probably we can make the mean and median meets.

We create a function `na_fill(data, pervalue)` which the `data` arguement is the `rdata` data frame and `pervalue` arguement is the `steps_per_interval` data frame.

```{r fill_na}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
```

We check that are there any missing values remaining or not

```{r check_empty}
sum(is.na(rdata_fill$steps))
```

Zero output shows that there are ***NO MISSING VALUES***.


### A histogram of the total number of steps taken each day

```{r histo_fill}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 

```

### Calculate and report the **mean** and **median** total number of steps taken per day.

```{r meanmedian_fill}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

The mean is **`r format(steps_mean_fill,digits = 8)`** and median is **`r format(steps_median_fill,digits = 8)`**.

### Do these values differ from the estimates from the first part of the assignment?

Yes, these values do differ slightly.

- **Before filling the data**
    1. Mean  : **`r format(steps_mean,digits = 8)`**
    2. Median: **`r format(steps_median,digits = 8)`**
    
    
- **After filling the data**
    1. Mean  : **`r format(steps_mean_fill,digits = 8)`**
    2. Median: **`r format(steps_median_fill,digits = 8)`**

We see that the values after filling the data mean and median are equal.

### What is the impact of inputing missing data on the estimates of the total daily number of steps?

After inputing the missing data, the mean value remains unchanged and the median value has shifted and almost matches to the mean.  

The data has shown a t-student distribution (see histograms) and it seems that the impact of imputing missing values has increase our peak, but it's not affect negatively our predictions. 


## Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.  
1. Augment the table with a column that indicates the day of the week  
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).  
3. Tabulate the average steps per interval for each data set.  
4. Plot the two data sets side by side for comparison.  

```{r weekdays}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
    
    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)
    
    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
    
    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```

Below you can see the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:
```{r plot_weekdays}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()

```

We can see at the graph above that activity on the weekday has the greatest peak from all steps intervals. But, we can see too that weekends activities has more peaks over a hundred than weekday. This could be due to the fact that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport. In the other hand, at weekend we can see better distribution of effort along the time.
